---
title: "NFL database"
output: html_document
date: "2023-09-21"
---
During the scientific programming project I will do a statistical analysis on the performance of Quarterbacks(QB) in the NFL. I will look at what effect age has on performance and if there are significant differences in QB-play between generation. Furthermore, I will look at what stats across the entire offense are most important in determining how productive the offense is. Lastly, I will also create and evaluate a function that improves the current rating used to evaluate QB's. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r import, include=FALSE}
#The following commands set the directory and load in all the data
setwd("C:/Users/MK122/OneDrive/Bureaublad/School/Systems Biology/Scientific programming/NFL database")
Quarterback_gamelogs = read.csv("Game_Logs_Quarterback.csv")
Runningback_gamelogs = read.csv('Game_Logs_Runningback.csv')
WideReceiver_gamelogs = read.csv("Game_Logs_Wide_Receiver_and_Tight_End.csv")
Kicker_gamelogs = read.csv("Game_Logs_Kickers.csv")
Basic_stats = read.csv("Basic_stats.csv")

#Most data in the data frame is not set to numeric so we change that where needed
Numeric_Columns_QB_Gamelogs = c(4,6,12:28) 
Quarterback_gamelogs[Numeric_Columns_QB_Gamelogs] = lapply(Quarterback_gamelogs[Numeric_Columns_QB_Gamelogs], as.numeric)
Numeric_Columns_RB_Gamelogs = c(4,6,12:25)
Runningback_gamelogs[Numeric_Columns_RB_Gamelogs] = lapply(Runningback_gamelogs[Numeric_Columns_RB_Gamelogs], as.numeric)
Numeric_Columns_WR_Gamelogs = c(4,6,12:25)
WideReceiver_gamelogs[Numeric_Columns_RB_Gamelogs] = lapply(WideReceiver_gamelogs[Numeric_Columns_WR_Gamelogs], as.numeric)
Numeric_Columns_K_Gamelogs = c(4,6,12:27)
Kicker_gamelogs[Numeric_Columns_RB_Gamelogs] = lapply(Kicker_gamelogs[Numeric_Columns_K_Gamelogs], as.numeric)
```

```{r age}
#The following 2 lines of code translate the birthday of players into a numeric birth year which I can use to calculate their age during each game
Basic_stats$Birthyear = substring(Basic_stats$Birthday, regexpr("/", Basic_stats$Birthday) + 1)
Basic_stats$Birthyear = as.numeric(substring(Basic_stats$Birthyear, regexpr("/", Basic_stats$Birthyear) + 1))

#The following code adds the birth year to each gamelog for all offensive players by using the Player.ID. The age is then calculated for each game
Quarterback_gamelogs = left_join(Quarterback_gamelogs, Basic_stats %>% select(Player.Id, Birthyear), by = "Player.Id")
Quarterback_gamelogs$Age =  Quarterback_gamelogs$Year - Quarterback_gamelogs$Birthyear

Runningback_gamelogs = left_join(Runningback_gamelogs, Basic_stats %>% select(Player.Id, Birthyear), by = "Player.Id")
Runningback_gamelogs$Age =  Runningback_gamelogs$Year - Runningback_gamelogs$Birthyear

WideReceiver_gamelogs = left_join(WideReceiver_gamelogs, Basic_stats %>% select(Player.Id, Birthyear), by = "Player.Id")
WideReceiver_gamelogs$Age =  WideReceiver_gamelogs$Year - WideReceiver_gamelogs$Birthyear

Kicker_gamelogs = left_join(Kicker_gamelogs, Basic_stats %>% select(Player.Id, Birthyear), by = "Player.Id")
Kicker_gamelogs$Age =  Kicker_gamelogs$Year - Kicker_gamelogs$Birthyear
```

```{r Game_inclusion, warning=FALSE}
#Removes data of games where the players didn't actually play or games outside of the regular season which aren't representative of how they play
Quarterback_gamelogs = filter(Quarterback_gamelogs, Passes.Attempted>0)
Quarterback_gamelogs = filter(Quarterback_gamelogs, Season=="Regular Season")
Quarterback_gamelogs$Offensive.Points = as.numeric(sub(" .*", "", Quarterback_gamelogs$Score))

#Runningback
Runningback_gamelogs = filter(Runningback_gamelogs, Rushing.Attempts>0)
Runningback_gamelogs = filter(Runningback_gamelogs, Season=="Regular Season")

#WideReceiver
WideReceiver_gamelogs = filter(WideReceiver_gamelogs, Receptions>0)
WideReceiver_gamelogs = filter(WideReceiver_gamelogs, Season=="Regular Season")

#Kicker
Kicker_gamelogs = filter(Kicker_gamelogs, FGs.Attempted>0 | Extra.Points.Attempted>0)
Kicker_gamelogs = filter(Kicker_gamelogs, Season=="Regular Season")
```


```{r Data Manipulation, warning=FALSE, include=FALSE}
#Changes NA's into 0 so we can sum them later on
Quarterback_gamelogs = Quarterback_gamelogs %>% mutate_all(~ifelse(is.na(.), 0, .))
Runningback_gamelogs = Runningback_gamelogs %>% mutate_all(~ifelse(is.na(.), 0, .))
WideReceiver_gamelogs = WideReceiver_gamelogs %>% mutate_all(~ifelse(is.na(.), 0, .))
Kicker_gamelogs = Kicker_gamelogs %>% mutate_all(~ifelse(is.na(.), 0, .))

#This specifies which columns I want to sum for each type of player
Sum_Columns_QB = c(14,15,17,19:24,26:28)
Column_names_QB = colnames(Quarterback_gamelogs[Sum_Columns_QB])

Sum_Columns_RB = c(14,15,17:20,22:25)
Column_names_RB = colnames(Runningback_gamelogs[Sum_Columns_RB])

Sum_Columns_WR = c(14,15,17:20,22:25)
Column_names_WR = colnames(WideReceiver_gamelogs[Sum_Columns_WR])

Sum_Columns_K = c(14:17,19,20)
Column_names_K = colnames(Kicker_gamelogs[Sum_Columns_K])

#Not every game is played by only one player in each position. The following code combines QB's that played the same opponent on the same week of the same year, which guarantees that the both are part of the same team that played that day. It then combines the names of the QB's and sums their stats across the game. I make a new data frame for this because if I would overwrite the original gamelogs I wouldn't have the individual stats of players in a game anymore
Offensive_gamelog_QB = Quarterback_gamelogs %>%
  group_by(Year, Week, Opponent, Outcome, Home.or.Away, Score, Offensive.Points) %>%
  summarize(
    Player.ID = paste(Player.Id, collapse = " + "),
    across(Column_names_QB,sum),
  ) %>%
  ungroup()

#The following code adds QB_ to the column names. 
Column_names_QB = paste("QB_", names(Offensive_gamelog_QB)[8:20], sep = "")
names(Offensive_gamelog_QB)[8:20] = Column_names_QB

#Does the same for Runningback
Offensive_gamelog_RB = Runningback_gamelogs %>%
  group_by(Year, Week, Opponent) %>%
  summarize(
    Player.ID = paste(Player.Id, collapse = " + "),
    across(Column_names_RB,sum),
  ) %>%
  ungroup()
Column_names_RB = paste("RB_", names(Offensive_gamelog_RB)[4:14], sep = "")
names(Offensive_gamelog_RB)[4:14] = Column_names_RB

#Does the same for Wide Receivers
Offensive_gamelog_WR = WideReceiver_gamelogs %>%
  group_by(Year, Week, Opponent) %>%
  summarize(
    Player.ID = paste(Player.Id, collapse = " + "),
    across(Column_names_WR,sum),
  ) %>%
  ungroup()
Column_names_WR = paste("WR_", names(Offensive_gamelog_WR)[4:14], sep = "")
names(Offensive_gamelog_WR)[4:14] = Column_names_WR

#Does the same for Kickers
Offensive_gamelog_K = Kicker_gamelogs %>%
  group_by(Year, Week, Opponent) %>%
  summarize(
    Player.ID = paste(Player.Id, collapse = " + "),
    across(Column_names_K,sum),
  ) %>%
  ungroup()
Column_names_K = paste("K_", names(Offensive_gamelog_K)[4:10], sep = "")
names(Offensive_gamelog_K)[4:10] = Column_names_K

#Combining the gamelogs of all offensive players into 1 combined data frame
Offensive_gamelog_Combined = left_join(Offensive_gamelog_QB, Offensive_gamelog_RB %>% select(RB_Player.ID,Column_names_RB,Year,Week,Opponent), by = c("Week", "Year", "Opponent"))
Offensive_gamelog_Combined = left_join(Offensive_gamelog_Combined, Offensive_gamelog_WR %>% select(WR_Player.ID,Column_names_WR,Year,Week,Opponent), by = c("Week", "Year", "Opponent"))
Offensive_gamelog_Combined = left_join(Offensive_gamelog_Combined, Offensive_gamelog_K %>% select(K_Player.ID,Column_names_K,Year,Week,Opponent), by = c("Week", "Year", "Opponent"))
Offensive_gamelog_Combined = Offensive_gamelog_Combined %>%
  mutate_if(~!is.list(.), ~ifelse(is.na(.), 0, .))
```
During a down, a lot of things can happen. QB can typically do one of 3 things. Throw the ball to Wide Receiver(WR) to attempt a pass, give the ball to the Runningback(RB) who will try to rush the ball forward, or rush the ball themselves. When the ball is thrown it can be intercepted by the defense of the opposing team. Furthermore whenever the QB gets tackled before getting the ball out it is called a sack. Lastly when any player on the field looses the ball it is called a fumble which can be recovered by both the defense and the offense. Based on these numbers, 


The following code creates a passer rating. This is a score developed by the NFL to evaluate how well a QB plays during a game. It 
```{r Passer rating}
#The following code creates a passer rating. This is a score developed by the NFL to evaluate how well a QB played during a game. It 
#Create original passer rating function
calculate_passer_rating <- function(COMP, ATT, YDS, TD, INT) {
  
  # Calculate the four components of the passer rating
  a = ((COMP / ATT) - 0.3) * 5
  b = ((YDS / ATT) - 3) * 0.25
  c = (TD / ATT) * 20
  d = 2.375 - ((INT / ATT) * 25)
  
  # Ensure that a,b,c&d are below 2.375
  a = pmin(a, 2.375)
  a = pmax(a, 0)
  b = pmin(b, 2.375)
  b = pmax(b,0)
  c = pmin(c, 2.375)
  c = pmax(c,0)
  d = pmax(d, 0)
  
  # Calculate the passer rating
  passer_rating = ((a+b+c+d)/6) * 100

  
  return(passer_rating)
}

Quarterback_gamelogs$Passer.Rating = calculate_passer_rating(Quarterback_gamelogs$Passes.Completed,Quarterback_gamelogs$Passes.Attempted, Quarterback_gamelogs$Passing.Yards, Quarterback_gamelogs$TD.Passes, Quarterback_gamelogs$Ints)
Offensive_gamelog_Combined$QB_Passer.Rating = calculate_passer_rating(Offensive_gamelog_Combined$QB_Passes.Completed,Offensive_gamelog_Combined$QB_Passes.Attempted, Offensive_gamelog_Combined$QB_Passing.Yards, Offensive_gamelog_Combined$QB_TD.Passes, Offensive_gamelog_Combined$QB_Ints)

```


```{r Passer rating2.0}
calculate_passer_rating = function(COMP, ATT, YDS, TD, INT, SACK, S.YDS, RUSH, R.YDS, FUMB, FUMB.L) {
  
  # Calculate the seven components of the passer rating
  a = ((COMP / ATT) - 0.3) * 5
  b = ((YDS / ATT) - 3) * 0.25
  c = (TD / ATT) * 20
  d = 2.375 - ((INT / ATT) * 25)
  e = ifelse(SACK == 0, 1, 1.000 - ((SACK / ATT) * ((S.YDS / SACK) + 3)))
  f = ifelse(RUSH == 0, 0, ((R.YDS / RUSH) - 3) * 0.25)
  g = 2.375 - ((FUMB/ATT)*3 + (FUMB.L/ATT)*22)
  
  # Ensure that a,b,c&d are below 2.375
  a = pmin(a, 2.375)
  a = pmax(a, 0)
  b = pmin(b, 2.375)
  b = pmax(b,0)
  c = pmin(c, 2.375)
  c = pmax(c,0)
  d = pmax(d, 0)
  e = pmax(e, 0)
  f = pmin(f,1)
  f = pmax(f,0)
  g = pmax(g,0)
  
  # Calculate the passer rating
  passer_rating = ((a + b + c + d + e + f + g) / 8.4) * 100
  
  return(passer_rating)
}

Quarterback_gamelogs$Passer.Rating = calculate_passer_rating(Quarterback_gamelogs$Passes.Completed,Quarterback_gamelogs$Passes.Attempted, Quarterback_gamelogs$Passing.Yards, Quarterback_gamelogs$TD.Passes, Quarterback_gamelogs$Ints, Quarterback_gamelogs$Sacks, Quarterback_gamelogs$Sacked.Yards.Lost, Quarterback_gamelogs$Rushing.Attempts, Quarterback_gamelogs$Rushing.Yards, Quarterback_gamelogs$Fumbles, Quarterback_gamelogs$Fumbles.Lost)

```



```{r Best Age Quarterback, warning=FALSE}
#The following chooses the stats that we want to compare across ages 
Age_stats_Columns_QB = c(15,17,20,21,23,31,32)
Age_stats_Columns_QB = colnames(Quarterback_gamelogs[Age_stats_Columns_QB])

#This code creates a dataframe in which the amount of average stats of each quarterback of a certain age is represented. Age groups with less than 100 games are excluded as that is a to small sample size
Age_stats_QB = Quarterback_gamelogs %>%
  group_by(Age) %>%
  summarize(
    Games = n(),
    across(Age_stats_Columns_QB,mean),
  ) %>%
  ungroup() %>%
  filter(Games >= 100)

#Create a plot showing the Ints, Sacks, Rushing attempts, Passes attempted and Offensive points
plot = ggplot(data = Age_stats_QB, aes(x = Age)) +
  geom_line(aes(y = Ints, color = "Ints"), linetype = "solid") +
  geom_line(aes(y = Sacks, color = "Sacks"), linetype = "solid") +
  geom_line(aes(y = Rushing.Attempts, color = "Rushing Attempts"), linetype = "solid") +
  geom_line(aes(y = Passes.Attempted / 10, color = "Passes Attempted"), linetype = "solid") +
  geom_line(aes(y = Offensive.Points / 10, color = "Offensive Points"), linetype = "solid") +
  labs(x = "Age", title = "Quarterback Statistics")

#Adds a y-axis on both the left and the right. The left Y-axis corresponds with the amount of Ints, Sacks and Rushing attempts. The right corresponds with the passes attempted and the offensive points.
plot = plot + scale_y_continuous(
  name = "Amount of Ints, Sacks & Rushing attempts",
  sec.axis = sec_axis(trans = ~ . * 10, name = "Amount of Offensive points & Passes attempted"),
  breaks = seq(0.5, 3.5, by = 0.5),
  limits = c(0.5, 3.5)
)

#Adds the x-axis and shows the amount of games played on each age
plot = plot + scale_x_continuous(breaks = Age_stats_QB$Age[seq(1, nrow(Age_stats_QB), by = 4)], 
                                  labels = paste(Age_stats_QB$Age[seq(1, nrow(Age_stats_QB), by = 4)], "(", Age_stats_QB$Games[seq(1, nrow(Age_stats_QB), by = 5)], " games)"))

#Creates a functioning legend
custom_colors = c("Ints" = "blue", "Rushing Attempts" = "red", "Sacks" = "green", "Passes Attempted" = "purple", "Offensive Points" = "orange")
plot = plot + scale_color_manual(values = custom_colors)

plot = plot + theme(legend.position="top")
# Print the plot
print(plot)
```


```{r Regression best age QB}
regression_results <- data.frame()
# Loop through variable pairs
rows_to_regress = c(3, 5:8)
for (i in rows_to_regress) {  
  DV <- names(Age_stats_QB)[i]  
  IV <- "Age"  
  
  # Fit the regression model
  model <- lm(paste(DV, "~", IV), data = Age_stats_QB)
  
  # Extract relevant information
  intercept <- coef(model)[1]
  coefficient <- coef(model)[2]
  p_value <- summary(model)$coefficients[2, 4]
  adj_r_2 <- summary(model)$adj.r.squared
  
  # Add results to the dataframe
  result_row <- data.frame(DV, IV, intercept, coefficient, p_value, adj_r_2)
  regression_results <- rbind(regression_results, result_row)
}

# Print or analyze the regression results
print(regression_results)
```

What we can gather from this data is that aging quarterbacks typically tend to get better over time creating more offensive output per game. They do this by an increase of their passing attempts and a decrease in their rushing attempts. This can be explained by the risk of injury that rushing attempts have which does not suit an older QB

```{r Career Stats}
#Creates the player stats
Career_stats = Quarterback_gamelogs %>%
  group_by(Name) %>%
  summarise(
    Wins = sum(Outcome == "W"),
    Amount_of_games = n(),
    Passing.Yards = sum(Passing.Yards),
    Passes.Completed = sum(Passes.Completed),
    Passes.Attempted = sum(Passes.Attempted),
    Sacks = sum(Sacks),
    Sacked.Yards.Lost = sum(Sacked.Yards.Lost),
    Rushing.Attempts = sum(Rushing.Attempts),
    Rushing.Yards = sum(Rushing.Yards),
    Fumbles = sum(Fumbles),
    Fumbles.Lost = sum(Fumbles.Lost),
    Ints = sum(Ints),
    TD.Passes = sum(TD.Passes),
    Offensive.Points = mean(Offensive.Points)
  ) %>%
  filter(Amount_of_games>=5)

Career_stats$Wins[is.na(Career_stats$Wins)] = 0
Career_stats$WinPercentage = floor(Career_stats$Wins/Career_stats$Amount_of_games *100)
Career_stats$Passer.Rating = calculate_passer_rating(Career_stats$Passes.Completed,Career_stats$Passes.Attempted, Career_stats$Passing.Yards, Career_stats$TD.Passes, Career_stats$Ints, Career_stats$Sacks, Career_stats$Sacked.Yards.Lost, Career_stats$Rushing.Attempts, Career_stats$Rushing.Yards, Career_stats$Fumbles, Career_stats$Fumbles.Lost)

Career_stats$games_group = cut(Career_stats$Amount_of_games, 
                        breaks = c(-Inf, 10, 150, Inf),
                        labels = c("Below 10", "10-150", "Above 150"))
color_mapping = c("Below 10" = "black", "10-150" = "blue", "Above 150" = "red")

ggplot(Career_stats, aes(x = WinPercentage, y = Passer.Rating, color = games_group)) +
  geom_point() +
  scale_color_manual(values = color_mapping) +
  labs(x = "WinPercentage", y = "Passer.Rating") +
  ggtitle("Scatter Plot of Passer Rating vs. Win Percentage")
```

```{r Biplot}
# Divide the variables of interest by Amount_of_games
Career_stats_average <- Career_stats %>%
  mutate(
    Passing.Yards = Passing.Yards / Amount_of_games,
    Passes.Completed = Passes.Completed / Amount_of_games,
    Passes.Attempted = Passes.Attempted / Amount_of_games,
    Sacks = Sacks / Amount_of_games,
    Sacked.Yards.Lost = Sacked.Yards.Lost / Amount_of_games,
    Rushing.Attempts = Rushing.Attempts / Amount_of_games,
    Rushing.Yards = Rushing.Yards / Amount_of_games,
    Fumbles = Fumbles / Amount_of_games,
    Fumbles.Lost = Fumbles.Lost / Amount_of_games,
    Ints = Ints / Amount_of_games,
    TD.Passes = TD.Passes / Amount_of_games,
    Offensive.Points = Offensive.Points / Amount_of_games
  )




# Select the numeric variables for PCA (excluding non-numeric and response variables)
variables_of_interest <- Career_stats_average[, c(
 "Passes.Completed", "Passes.Attempted",
  "Sacks", "Rushing.Attempts", "Fumbles", "Ints", "TD.Passes",
  "Offensive.Points"
)]

# Standardize the data (mean center and scale)
standardized_data <- scale(variables_of_interest)

# Perform PCA
pca_result <- prcomp(standardized_data)

# Summary of PCA results
summary(pca_result)

# Variance explained by each principal component
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)

# Scree plot to visualize variance explained
scree_plot <- ggplot() +
  geom_bar(aes(x = 1:length(var_explained), y = var_explained), stat = "identity") +
  labs(x = "Principal Component", y = "Variance Explained") +
  ggtitle("Scree Plot")

# Plot of the first two principal components
pc1_pc2_plot <- ggplot(as.data.frame(pca_result$x), aes(x = PC1, y = PC2)) +
  geom_point() +
  labs(x = "PC1", y = "PC2") +
  ggtitle("PCA: PC1 vs. PC2")

# Print or visualize the PCA results and plots
print(pca_result)
print(scree_plot)
print(pc1_pc2_plot)

# Create a biplot
biplot(pca_result, scale = 0, col = c("white", "black"))
```

