---
title: "NFL database"
output: html_document
date: "2023-09-21"
---
During the scientific programming project I will do a statistical analysis on the performance of Quarterbacks(QB) in american football in the NFL(official national league of the US). I will look at what effect age has on performance and if there are significant differences in QB-play between generation. Furthermore, I will look at what stats across the entire offense are most important in determining how productive the offense is. Lastly, I will also create and evaluate a function that improves the current rating used to evaluate QB's. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

```{r packages}
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)
```

```{r import}
#The following commands set the directory and load in all the data
setwd("C:/Users/MK122/OneDrive/Bureaublad/School/Systems Biology/Scientific programming/Final skills session/Git/Scientific-programming-project-Melle")
Quarterback_gamelogs = read.csv("Game_Logs_Quarterback.csv")
Runningback_gamelogs = read.csv('Game_Logs_Runningback.csv')
WideReceiver_gamelogs = read.csv("Game_Logs_Wide_Receiver_and_Tight_End.csv")
Kicker_gamelogs = read.csv("Game_Logs_Kickers.csv")
Basic_stats = read.csv("Basic_stats.csv")

#Most data in the data frame is not set to numeric so we change that where needed
Numeric_Columns_QB_Gamelogs = c(4,6,12:28) 
Quarterback_gamelogs[Numeric_Columns_QB_Gamelogs] = lapply(Quarterback_gamelogs[Numeric_Columns_QB_Gamelogs], as.numeric)
Numeric_Columns_RB_Gamelogs = c(4,6,12:25)
Runningback_gamelogs[Numeric_Columns_RB_Gamelogs] = lapply(Runningback_gamelogs[Numeric_Columns_RB_Gamelogs], as.numeric)
Numeric_Columns_WR_Gamelogs = c(4,6,12:25)
WideReceiver_gamelogs[Numeric_Columns_RB_Gamelogs] = lapply(WideReceiver_gamelogs[Numeric_Columns_WR_Gamelogs], as.numeric)
Numeric_Columns_K_Gamelogs = c(4,6,12:27)
Kicker_gamelogs[Numeric_Columns_RB_Gamelogs] = lapply(Kicker_gamelogs[Numeric_Columns_K_Gamelogs], as.numeric)
```

```{r age}
#The following 2 lines of code translate the birthday of players into a numeric birth year which I can use to calculate their age during each game
Basic_stats$Birthyear = substring(Basic_stats$Birthday, regexpr("/", Basic_stats$Birthday) + 1)
Basic_stats$Birthyear = as.numeric(substring(Basic_stats$Birthyear, regexpr("/", Basic_stats$Birthyear) + 1))

#The following code adds the birth year to each gamelog for all offensive players by using the Player.ID. The age is then calculated for each game
Quarterback_gamelogs = left_join(Quarterback_gamelogs, Basic_stats %>% select(Player.Id, Birthyear), by = "Player.Id")
Quarterback_gamelogs$Age =  Quarterback_gamelogs$Year - Quarterback_gamelogs$Birthyear

Runningback_gamelogs = left_join(Runningback_gamelogs, Basic_stats %>% select(Player.Id, Birthyear), by = "Player.Id")
Runningback_gamelogs$Age =  Runningback_gamelogs$Year - Runningback_gamelogs$Birthyear

WideReceiver_gamelogs = left_join(WideReceiver_gamelogs, Basic_stats %>% select(Player.Id, Birthyear), by = "Player.Id")
WideReceiver_gamelogs$Age =  WideReceiver_gamelogs$Year - WideReceiver_gamelogs$Birthyear

Kicker_gamelogs = left_join(Kicker_gamelogs, Basic_stats %>% select(Player.Id, Birthyear), by = "Player.Id")
Kicker_gamelogs$Age =  Kicker_gamelogs$Year - Kicker_gamelogs$Birthyear
```

```{r Game_inclusion}
#Removes data of games where the players didn't actually play or games outside of the regular season which aren't representative of how they play
Quarterback_gamelogs = filter(Quarterback_gamelogs, Passes.Attempted>0)
Quarterback_gamelogs = filter(Quarterback_gamelogs, Season=="Regular Season")
Quarterback_gamelogs$Offensive.Points = as.numeric(sub(" .*", "", Quarterback_gamelogs$Score))

#Runningback
Runningback_gamelogs = filter(Runningback_gamelogs, Rushing.Attempts>0)
Runningback_gamelogs = filter(Runningback_gamelogs, Season=="Regular Season")

#WideReceiver
WideReceiver_gamelogs = filter(WideReceiver_gamelogs, Receptions>0)
WideReceiver_gamelogs = filter(WideReceiver_gamelogs, Season=="Regular Season")

#Kicker
Kicker_gamelogs = filter(Kicker_gamelogs, FGs.Attempted>0 | Extra.Points.Attempted>0)
Kicker_gamelogs = filter(Kicker_gamelogs, Season=="Regular Season")
```


```{r Data Manipulation}
#Changes NA's into 0 so we can sum them later on
Quarterback_gamelogs = Quarterback_gamelogs %>% mutate_all(~ifelse(is.na(.), 0, .))
Runningback_gamelogs = Runningback_gamelogs %>% mutate_all(~ifelse(is.na(.), 0, .))
WideReceiver_gamelogs = WideReceiver_gamelogs %>% mutate_all(~ifelse(is.na(.), 0, .))
Kicker_gamelogs = Kicker_gamelogs %>% mutate_all(~ifelse(is.na(.), 0, .))

#This specifies which columns I want to sum for each type of player
Sum_Columns_QB = c(14,15,17,19:24,26:28)
Column_names_QB = colnames(Quarterback_gamelogs[Sum_Columns_QB])

Sum_Columns_RB = c(14,15,17:20,22:25)
Column_names_RB = colnames(Runningback_gamelogs[Sum_Columns_RB])

Sum_Columns_WR = c(14,15,17:20,22:25)
Column_names_WR = colnames(WideReceiver_gamelogs[Sum_Columns_WR])

Sum_Columns_K = c(14:17,19,20)
Column_names_K = colnames(Kicker_gamelogs[Sum_Columns_K])

#Not every game is played by only one player in each position. The following code combines QB's that played the same opponent on the same week of the same year, which guarantees that the both are part of the same team that played that day. It then combines the names of the QB's and sums their stats across the game. I make a new data frame for this because if I would overwrite the original gamelogs I wouldn't have the individual stats of players in a game anymore
Offensive_gamelog_QB = Quarterback_gamelogs %>%
  group_by(Year, Week, Opponent, Outcome, Home.or.Away, Score, Offensive.Points) %>%
  summarize(
    Player.ID = paste(Player.Id, collapse = " + "),
    across(Column_names_QB,sum),
    Age = mean(Age),
  ) %>%
  ungroup()

#The following code adds QB_ to the column names. 
Column_names_QB = paste("QB_", names(Offensive_gamelog_QB)[8:21], sep = "")
names(Offensive_gamelog_QB)[8:21] = Column_names_QB

#Does the same for Runningback
Offensive_gamelog_RB = Runningback_gamelogs %>%
  group_by(Year, Week, Opponent) %>%
  summarize(
    Player.ID = paste(Player.Id, collapse = " + "),
    across(Column_names_RB,sum),
  ) %>%
  ungroup()
Column_names_RB = paste("RB_", names(Offensive_gamelog_RB)[4:14], sep = "")
names(Offensive_gamelog_RB)[4:14] = Column_names_RB

#Does the same for Wide Receivers
Offensive_gamelog_WR = WideReceiver_gamelogs %>%
  group_by(Year, Week, Opponent) %>%
  summarize(
    Player.ID = paste(Player.Id, collapse = " + "),
    across(Column_names_WR,sum),
  ) %>%
  ungroup()
Column_names_WR = paste("WR_", names(Offensive_gamelog_WR)[4:14], sep = "")
names(Offensive_gamelog_WR)[4:14] = Column_names_WR

#Does the same for Kickers
Offensive_gamelog_K = Kicker_gamelogs %>%
  group_by(Year, Week, Opponent) %>%
  summarize(
    Player.ID = paste(Player.Id, collapse = " + "),
    across(Column_names_K,sum),
  ) %>%
  ungroup()
Column_names_K = paste("K_", names(Offensive_gamelog_K)[4:10], sep = "")
names(Offensive_gamelog_K)[4:10] = Column_names_K

#Combining the gamelogs of all offensive players into 1 combined data frame
Offensive_gamelog_Combined = left_join(Offensive_gamelog_QB, Offensive_gamelog_RB %>% select(RB_Player.ID,Column_names_RB,Year,Week,Opponent), by = c("Week", "Year", "Opponent"))
Offensive_gamelog_Combined = left_join(Offensive_gamelog_Combined, Offensive_gamelog_WR %>% select(WR_Player.ID,Column_names_WR,Year,Week,Opponent), by = c("Week", "Year", "Opponent"))
Offensive_gamelog_Combined = left_join(Offensive_gamelog_Combined, Offensive_gamelog_K %>% select(K_Player.ID,Column_names_K,Year,Week,Opponent), by = c("Week", "Year", "Opponent"))
Offensive_gamelog_Combined = Offensive_gamelog_Combined %>%
  mutate_if(~!is.list(.), ~ifelse(is.na(.), 0, .))
```
During a down, a lot of things can happen. QB can typically do one of 3 things. Throw the ball to Wide Receiver(WR) to attempt a pass, give the ball to the Runningback(RB) who will try to rush the ball forward, or rush the ball themselves. When the ball is thrown it can be intercepted by the defense of the opposing team. Furthermore whenever the QB gets tackled before getting the ball out it is called a sack. Lastly when any player on the field looses the ball it is called a fumble which can be recovered by both the defense and the offense. Based on these numbers, a scoring for each game is made by the NFL of the QB performance called passer rating. 

The following code creates the passer rating. It incorporates successful passing percentage, yards per pass, touchdowns per pass and interceptions per pass. Scores are between 0 and 158.3 
```{r Passer rating, echo=TRUE}
#Create original passer rating function
calculate_passer_rating = function(COMP, ATT, YDS, TD, INT) {
  
  # Calculate the four components of the passer rating
  a = ((COMP / ATT) - 0.3) * 5
  b = ((YDS / ATT) - 3) * 0.25
  c = (TD / ATT) * 20
  d = 2.375 - ((INT / ATT) * 25)
  
  # Ensure that a,b,c&d are below 2.375 and above 0
  a = pmin(a, 2.375)
  a = pmax(a, 0)
  b = pmin(b, 2.375)
  b = pmax(b,0)
  c = pmin(c, 2.375)
  c = pmax(c,0)
  d = pmax(d, 0)
  
  # Calculate the passer rating
  passer_rating = ((a+b+c+d)/6) * 100

  
  return(passer_rating)
}
```

Since Passer rating does not include the entire skillset of a QB I have extended the function to include rushing, sacks and fumbles creating Passer rating2.0. Scores are more centered and are between 0 and 158.3 due to the fact that there are more variables that have to be minimized or maximized to create polarizing scores.
```{r Passer rating2.0, echo=TRUE}
#Create an improved passer rating function
calculate_passer_rating2.0 = function(COMP, ATT, YDS, TD, INT, SACK, S.YDS, RUSH, R.YDS, FUMB, FUMB.L) {
  # Calculate the seven components of the passer rating2.0
  a = ((COMP / ATT) - 0.3) * 5
  b = ((YDS / ATT) - 3) * 0.25
  c = (TD / ATT) * 20
  d = 2.375 - ((INT / ATT) * 25)
  e = ifelse(SACK == 0, 1.1875, 1.1875 - ((SACK / ATT) * (S.YDS / SACK)))
  f = ifelse(RUSH == 0, 0, ((R.YDS / RUSH) - 3) * 0.25)
  g = 2.375 - ((FUMB/ATT)*3 + (FUMB.L/ATT)*22)
  
  # Ensure that a,b,c & d are below 2.375, f is below 1.18875 and that everything is above 0
  a = pmin(a, 2.375)
  a = pmax(a, 0)
  b = pmin(b, 2.375)
  b = pmax(b,0)
  c = pmin(c, 2.375)
  c = pmax(c,0)
  d = pmax(d, 0)
  e = pmax(e, 0)
  f = pmin(f,1.1875)
  f = pmax(f,0)
  g = pmax(g,0)
  
  # Calculate the passer rating2.0
  passer_rating2.0 = ((a + b + c + d + e + f + g) / 9) * 100
  
  return(passer_rating2.0)
}
```

Underneath is a barchart that shows the distribution of passer rating scores

```{r  passer_rating calculation and bar chart}
# Calculates the passer rating for each QB in a game
Quarterback_gamelogs$Passer.Rating = calculate_passer_rating(Quarterback_gamelogs$Passes.Completed,Quarterback_gamelogs$Passes.Attempted, Quarterback_gamelogs$Passing.Yards, Quarterback_gamelogs$TD.Passes, Quarterback_gamelogs$Ints)

# Calculates the Passer rating for all plays made by a QB in a game
Offensive_gamelog_Combined$QB_Passer.Rating = calculate_passer_rating(Offensive_gamelog_Combined$QB_Passes.Completed,Offensive_gamelog_Combined$QB_Passes.Attempted, Offensive_gamelog_Combined$QB_Passing.Yards, Offensive_gamelog_Combined$QB_TD.Passes, Offensive_gamelog_Combined$QB_Ints)

# Calculates the passer rating 2.0 for each QB in a game
Quarterback_gamelogs$Passer.Rating2.0 = calculate_passer_rating2.0(Quarterback_gamelogs$Passes.Completed,Quarterback_gamelogs$Passes.Attempted, Quarterback_gamelogs$Passing.Yards, Quarterback_gamelogs$TD.Passes, Quarterback_gamelogs$Ints, Quarterback_gamelogs$Sacks, Quarterback_gamelogs$Sacked.Yards.Lost, Quarterback_gamelogs$Rushing.Attempts, Quarterback_gamelogs$Rushing.Yards, Quarterback_gamelogs$Fumbles, Quarterback_gamelogs$Fumbles.Lost)

#Calculates the Passer rating 2.0 for all plays made by a QB in a game
Offensive_gamelog_Combined$QB_Passer.Rating2.0 = calculate_passer_rating2.0(Offensive_gamelog_Combined$QB_Passes.Completed,Offensive_gamelog_Combined$QB_Passes.Attempted, Offensive_gamelog_Combined$QB_Passing.Yards, Offensive_gamelog_Combined$QB_TD.Passes, Offensive_gamelog_Combined$QB_Ints, Offensive_gamelog_Combined$QB_Sacks, Offensive_gamelog_Combined$QB_Sacked.Yards.Lost, Offensive_gamelog_Combined$QB_Rushing.Attempts, Offensive_gamelog_Combined$QB_Rushing.Yards, Offensive_gamelog_Combined$QB_Fumbles, Offensive_gamelog_Combined$QB_Fumbles.Lost)

#Creating a bar chart

#Dividing the passer ratings into bins
bin_width = 5
bin_breaks = c(0, seq(5, 160, by = bin_width))

Quarterback_gamelogs$Passer.Rating = cut(Quarterback_gamelogs$Passer.Rating, 
                                            breaks = bin_breaks,
                                            right = FALSE)

Quarterback_gamelogs$Passer.Rating2.0 = cut(Quarterback_gamelogs$Passer.Rating2.0, 
                                              breaks = bin_breaks,
                                              right = FALSE)

# Create a bar chart with two bars for each bin
ggplot(Quarterback_gamelogs, aes(x = Passer.Rating)) +
  geom_bar(aes(fill = "Passer.Rating"), position = "dodge") +
  geom_bar(aes(x = Passer.Rating2.0, fill = "Passer.Rating2.0"), position = "dodge") +
  labs(x = "Passer Rating", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Passer.Rating" = "skyblue", "Passer.Rating2.0" = "lightcoral")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


I perform an univariate regression on the full offensive performance to see which individual stat has the highest effect on offensive points. The results of the regression show which individual stat would be best to explain the variance in points scored by the offense. As expected, the Passer ratings have the highest scores with passer rating 2.0 slightly edging out the original passer rating
```{r Regression full offense univariate}
# Create the regression dataframe
regression_results_Offense = data.frame()
# Create a loop for all statistics in the offense that might have an effect on offensive production
Stats_of_Interest = c(9:21, 23:32, 34:43, 45:52)
for (i in Stats_of_Interest) {  
  DV = "Offensive.Points"    
  IV = names(Offensive_gamelog_Combined)[i] 
  
# Run the regression model
  model = lm(paste(DV, "~", IV), data = Offensive_gamelog_Combined)
  
# Extract relevant information
  intercept = coef(model)[1]
  coefficient = coef(model)[2]
  p_value = summary(model)$coefficients[2, 4]
  adj_r2 = summary(model)$adj.r.squared
  
# Add results to the dataframe
  result_row = data.frame(IV, coefficient, intercept, p_value, adj_r2)
  regression_results_Offense = rbind(regression_results_Offense, result_row)
}

# Round, order and print the regression results
regression_results_Offense$intercept = round(regression_results_Offense$intercept, digits = 3)
regression_results_Offense$coefficient = round(regression_results_Offense$coefficient, digits = 3)
regression_results_Offense$adj_r2 = round(regression_results_Offense$adj_r2, digits = 3)
regression_results_Offense = regression_results_Offense[order(-regression_results_Offense$adj_r2), ]
knitr::kable(regression_results_Offense, row.names = FALSE, format = "markdown")

# Create a bar chart of the 10 stats that have the biggest effect
ggplot(head(regression_results_Offense, 10), aes(x = reorder(IV, -adj_r2), y = adj_r2)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Variables", y = "Adj_r_squared", title = "What stat is the best predictor for offensive output") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Now I try the same but with a linear multivariate regression approach. This calculates how much each statistic individually explains of the variance in offensive points scored. Results show that althought the p-values are quite high, the variance explained by each individual variable is very low which might be caused by high correlation between variables.
```{r Regression full offense multivariate}
# Setting which variables are dependent and independent
dependent_variable = "Offensive.Points"
independent_variables = names(Offensive_gamelog_Combined)[Stats_of_Interest]

# Creating the formula that is going to be used for the regression
formula = as.formula(paste(dependent_variable, "~", paste(independent_variables, collapse = " + ")))

# Fit the regression model
model = lm(formula, data = Offensive_gamelog_Combined)

# Extract relevant information
coefficients <- coef(model)[-1] 
p_values <- summary(model)$coefficients[-1, 4]
intercept = summary(model)
adj_r2 <- summary(model)$adj.r.squared

# Create a summary dataframe
result_summary <- data.frame(
  IV = independent_variables,
  Coefficient = coefficients,
  P_Value = p_values,
  adj_r2 = adj_r2 
)

# Create a vector to store the part R-squared values
part_r_squared = numeric(length(independent_variables))

# Loop through the independent variables and calculate part R-squared
for (i in 1:length(independent_variables)) {
  IV_to_remove = independent_variables[i]
  
  # Fit the model without the current variable
  reduced_formula = as.formula(paste(dependent_variable, "~", paste(setdiff(independent_variables, IV_to_remove), collapse = " + ")))
  reduced_model = lm(reduced_formula, data = Offensive_gamelog_Combined)
  
  # Calculate the change in R-squared when the variable is removed
  delta_r_squared = adj_r2 - summary(reduced_model)$r.squared
  
  # Calculate the part R-squared for the current variable
  part_r_squared[i] = delta_r_squared
}

# Add part R-squared values to the result_summary dataframe
result_summary$Part_R_Squared = part_r_squared

# Round, order, and print the updated regression results
result_summary$Coefficient = round(result_summary$Coefficient, digits = 3)
result_summary$adj_r2 = round(result_summary$adj_r2, digits = 3)
result_summary$Part_R_Squared = round(result_summary$Part_R_Squared, digits = 3)
result_summary = result_summary[order(-result_summary$Part_R_Squared), ]

# Create a nice format for the table
knitr::kable(result_summary, row.names = FALSE, format = "markdown")

# Create a bar chart to visualize the part R-squared values
ggplot(head(result_summary, 10), aes(x = reorder(IV, -Part_R_Squared), y = Part_R_Squared)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Variables", y = "Part R-squared", title = "Individual Contribution of Variables to Offensive Output") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

To check whether the small values for part R_squared are caused by high correlation between statistics, a correlation matrix was made. This correlation matrix however did not produce extremely high results which could explain the values for part R_squared.
```{r Correlation matrix}

#redefining stats of interest to include only QB statistics
Stats_of_Interest = c(7,9:21,51,52)

#Saving the values for Stats of interest 
variables_of_interest = Offensive_gamelog_Combined[Stats_of_Interest]

# Calculate the correlation matrix
correlation_matrix = cor(variables_of_interest)

# Convert the correlation matrix to a format which can be put into a heatmap
correlation_data = melt(correlation_matrix)

# Create the correlation heatmap
ggplot(data = correlation_data, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "red", high = "blue", limits = c(-1, 1)) + 
  labs(title = "Correlation Heatmap",
       x = "QB statistics",
       y = "QB statistics") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Another question that I wanted to investigate was how age effects QB play. Therefore I've created a comparison of some of the most important variables of QB play.
```{r Best Age Quarterback}
#The following chooses the stats that I want to compare across ages 
Age_stats_Columns_QB = c(15,17,20,21,23,31)
Age_stats_Columns_QB = colnames(Quarterback_gamelogs[Age_stats_Columns_QB])

#This code creates a dataframe in which the amount of average stats of each quarterback of a certain age is represented. Age groups with less than 100 games are excluded as that is a to small sample size
Age_stats_QB = Quarterback_gamelogs %>%
  group_by(Age) %>%
  summarize(
    Games = n(),
    across(Age_stats_Columns_QB,mean),
  ) %>%
  ungroup() %>%
  filter(Games >= 100)

#Create a plot showing the Ints, Sacks, Rushing attempts, Passes attempted and Offensive points
plot = ggplot(data = Age_stats_QB, aes(x = Age)) +
  geom_line(aes(y = Ints, color = "Ints"), linetype = "solid") +
  geom_line(aes(y = Sacks, color = "Sacks"), linetype = "solid") +
  geom_line(aes(y = Rushing.Attempts, color = "Rushing Attempts"), linetype = "solid") +
  geom_line(aes(y = Passes.Attempted / 10, color = "Passes Attempted"), linetype = "solid") +
  geom_line(aes(y = Offensive.Points / 10, color = "Offensive Points"), linetype = "solid") +
  labs(x = "Age", title = "Effect of age on QB play")

#Adds a y-axis on both the left and the right. The left Y-axis corresponds with the amount of Ints, Sacks and Rushing attempts. The right corresponds with the passes attempted and the offensive points.
plot = plot + scale_y_continuous(
  name = "Amount of Ints, Sacks & Rushing attempts",
  sec.axis = sec_axis(trans = ~ . * 10, name = "Amount of Offensive points & Passes attempted"),
  breaks = seq(0.5, 3.5, by = 0.5),
  limits = c(0.5, 3.5)
)

#Adds the x-axis and shows the amount of games played on each age
plot = plot + scale_x_continuous(breaks = Age_stats_QB$Age[seq(1, nrow(Age_stats_QB), by = 4)], 
                                  labels = paste(Age_stats_QB$Age[seq(1, nrow(Age_stats_QB), by = 4)], "(", Age_stats_QB$Games[seq(1, nrow(Age_stats_QB), by = 5)], " games)"))

#Creates a legend
custom_colors = c("Ints" = "blue", "Rushing Attempts" = "red", "Sacks" = "green", "Passes Attempted" = "purple", "Offensive Points" = "orange")
plot = plot + scale_color_manual(values = custom_colors)

plot = plot + theme(legend.position="top")
# Print the plot
print(plot)
```


```{r Regression best age QB}
regression_results_Age = data.frame()
# Loop through variable pairs
rows_to_regress = c(7,10,13,14,16)
for (i in rows_to_regress) {  
  DV = names(Offensive_gamelog_QB)[i]  
  IV = "QB_Age"  
  
  # Fit the regression model
  model = lm(paste(DV, "~", IV), data = Offensive_gamelog_QB)
  
  # Extract relevant information
  intercept = coef(model)[1]
  coefficient = coef(model)[2]
  p_value = summary(model)$coefficients[2, 4]
  adj_r2 = summary(model)$adj.r.squared
  
  # Add results to the dataframe
  result_row = data.frame(DV, intercept, coefficient, p_value, adj_r2)
  regression_results_Age = rbind(regression_results_Age, result_row)
}

# Print, round and order the regression results
regression_results_Age$intercept = round(regression_results_Age$intercept, digits = 3)
regression_results_Age$coefficient = round(regression_results_Age$coefficient, digits = 3)
regression_results_Age$adj_r2 = round(regression_results_Age$adj_r2, digits = 3)
regression_results_Age = regression_results_Age[order(-regression_results_Age$adj_r2), ]
knitr::kable(regression_results_Age, row.names = FALSE, format = "markdown")
```

What we can gather from this data is that aging quarterbacks typically tend to get better over time creating more offensive output per game. They do this by an increase of their passing attempts and a decrease in their rushing attempts. This can be explained by the risk of injury that rushing attempts have which does not suit an older QB and the increase of experience that a older QB has to oversee the play and see free wide receivers

Furthermore, I was interested in which QB's have been the most successful over their entire career. Therefore, I've created the following scatter plot which compares passer rating to win percentage
```{r Career Stats}
#Creates the player stats over their entire career
Career_stats = Quarterback_gamelogs %>%
  group_by(Name, Birthyear) %>%
  summarise(
    Wins = sum(Outcome == "W"),
    Amount_of_games = n(),
    Passing.Yards = sum(Passing.Yards),
    Passes.Completed = sum(Passes.Completed),
    Passes.Attempted = sum(Passes.Attempted),
    Sacks = sum(Sacks),
    Sacked.Yards.Lost = sum(Sacked.Yards.Lost),
    Rushing.Attempts = sum(Rushing.Attempts),
    Rushing.Yards = sum(Rushing.Yards),
    Fumbles = sum(Fumbles),
    Fumbles.Lost = sum(Fumbles.Lost),
    Ints = sum(Ints),
    TD.Passes = sum(TD.Passes),
    Offensive.Points = mean(Offensive.Points)
  ) %>%
  filter(Amount_of_games>=5)

Career_stats$Wins[is.na(Career_stats$Wins)] = 0
Career_stats$WinPercentage = floor(Career_stats$Wins/Career_stats$Amount_of_games *100)
Career_stats$Passer.Rating = calculate_passer_rating(Career_stats$Passes.Completed,Career_stats$Passes.Attempted, Career_stats$Passing.Yards, Career_stats$TD.Passes, Career_stats$Ints)
Career_stats$Passer.Rating2.0 = calculate_passer_rating2.0(Career_stats$Passes.Completed,Career_stats$Passes.Attempted, Career_stats$Passing.Yards, Career_stats$TD.Passes, Career_stats$Ints, Career_stats$Sacks, Career_stats$Sacked.Yards.Lost, Career_stats$Rushing.Attempts, Career_stats$Rushing.Yards, Career_stats$Fumbles, Career_stats$Fumbles.Lost)

# Divides the players into 3 groups. Players that have less then 10 games, between 10 and 150 and above 150 games.
Career_stats$games_group = cut(Career_stats$Amount_of_games, 
                        breaks = c(-Inf, 10, 150, Inf),
                        labels = c("Below 10", "10-150", "Above 150"))
color_mapping = c("Below 10" = "black", "10-150" = "blue", "Above 150" = "red")

# Create the scatterplot
ggplot(Career_stats, aes(x = WinPercentage, y = Passer.Rating2.0, color = games_group)) +
  geom_point() +
  scale_color_manual(values = color_mapping) +
  labs(x = "WinPercentage", y = "Passer.Rating2.0") +
  ggtitle("Scatter Plot of Passer Rating2.0 vs. Win Percentage")

# Filter out the best QB's that have played a significant amount of games
filtered_qbs = Career_stats %>% filter(Passer.Rating2.0 > 90, WinPercentage > 60, Amount_of_games > 50)

# Select the columns of interest
result_table = filtered_qbs %>% select(Name, WinPercentage, Passer.Rating2.0, Amount_of_games) %>%
                                 arrange(desc(Passer.Rating2.0)) 

rownames(result_table) = NULL

# Print the result table
knitr::kable(result_table, format = "markdown")
```

In the scatterplot you can see that Quarterbacks that have played a lot of games then to be located in the top right, thus having a great career performance. This makes sense as better QB's tend to play more games and be replaced less often. This is also visible in the table with the top 10 QB's that is dominated by experienced QB's

Finally, I have performed a PCA to see if there are any characteristics that can clearly seperate QB's I have used the QB passer stats as well as sacks, offensive points, rumbles and rushing to base the PCA on. In the PCA the games are color coded for the generation in which the player has played, how many games they have played and their average passer rating.  

```{r PCA}
# Divide the variables of interest by Amount_of_games
Career_stats_average = Career_stats %>%
  mutate
    Passing.Yards = Passing.Yards / Amount_of_games,
    Passes.Completed = Passes.Completed / Amount_of_games,
    Passes.Attempted = Passes.Attempted / Amount_of_games,
    Sacks = Sacks / Amount_of_games,
    Sacked.Yards.Lost = Sacked.Yards.Lost / Amount_of_games,
    Rushing.Attempts = Rushing.Attempts / Amount_of_games,
    Rushing.Yards = Rushing.Yards / Amount_of_games,
    Fumbles = Fumbles / Amount_of_games,
    Fumbles.Lost = Fumbles.Lost / Amount_of_games,
    Ints = Ints / Amount_of_games,
    TD.Passes = TD.Passes / Amount_of_games,
    Generation = Birthyear + 28 #average age of a QB
  )


#Subdivide the players into their different generations
Career_stats_average$Generation_Category <- cut(
  Career_stats_average$Generation,
  breaks = c(1960, 1980, 2000, Inf),
  labels = c("1960-1980", "1980-2000", "> 2000")
)

#Subdivide the players into groups based on the amount of games played
Career_stats_average$Amount_of_games_Category = cut(
  Career_stats_average$Amount_of_games,
  breaks = c(-Inf, 10, 100, Inf),
  labels = c("Less then 10", "10-100", "> 100")
)

#subdivide the players based on their Passer Rating
Career_stats_average$Passer.Rating2.0_Category = cut(
  Career_stats_average$Passer.Rating2.0,
  breaks = c(-Inf, 70, 90, Inf),
  labels = c("Less then 70", "70-90", "> 90")
)

## Select the variables that are included into the PCA
variables_of_interest = Career_stats_average[, c(
  "Passes.Completed", "Passes.Attempted",
  "Sacks", "Rushing.Attempts", "Fumbles", "Ints", "TD.Passes",
  "Offensive.Points"
)]

# Standardize the data (mean center and scale)
standardized_data = scale(variables_of_interest)

# Perform PCA
pca_result = prcomp(standardized_data)

# Summary of PCA results
summary(pca_result)

# Variance explained by each principal component
var_explained = pca_result$sdev^2 / sum(pca_result$sdev^2)

# Scree plot to visualize variance explained by each PC
scree_plot = ggplot() +
  geom_bar(aes(x = 1:length(var_explained), y = var_explained), stat = "identity") +
  labs(x = "Principal Component", y = "Variance Explained") +
  ggtitle("Scree Plot")

#Create color mapping that is used for the generation separation
color_mapping = c(             
  "1960-1980" = "blue",         # Between 1960 and 1980
  "1980-2000" = "green",        # Between 1980 and 2000
  "> 2000" = "purple"           # Later than 2000
)

#Create color mapping that is used for the amount of games separation
color_mapping2 = c(             
  "Less then 10" = "blue",      # Less then 10
  "10-100" = "green",           # between 10 and 100
  "> 100" = "purple"            # More than 100
)

#Create color mapping that is used for the separation based on passer rating
color_mapping3 = c(             
  "Less then 70" = "blue",     # Less then 70
  "70-90" = "green",           # between 70 and 90
  "> 90" = "purple"            # More than 90
)

#create the data for the generation PCA
pca_data_generation = data.frame(PC1 = pca_result$x[, 1], PC2 = pca_result$x[, 2], Generation = Career_stats_average$Generation)


pc1_pc2_plot_Generation = ggplot(as.data.frame(pca_result$x), aes(x = PC1, y = PC2, color = Career_stats_average$Generation_Category)) +
  geom_point() +
  labs(x = "PC1", y = "PC2", color = "Generation") +
  ggtitle("PCA: PC1 vs. PC2 seperated by generation") +
  scale_color_manual(values = color_mapping)  



#create the data for the amount of games PCA
pca_data_Games = data.frame(PC1 = pca_result$x[, 1], PC2 = pca_result$x[, 2], Amount_of_games = Career_stats_average$Amount_of_games)


pc1_pc2_plot_Games = ggplot(as.data.frame(pca_result$x), aes(x = PC1, y = PC2, color = Career_stats_average$Amount_of_games_Category)) +
  geom_point() +
  labs(x = "PC1", y = "PC2", color = "Amount_of_games") +
  ggtitle("PCA: PC1 vs. PC2 seperated by amount of games") +
  scale_color_manual(values = color_mapping2)  


#create the data for the Passer rating PCA
pca_data_Passer.Rating2.0 = data.frame(PC1 = pca_result$x[, 1], PC2 = pca_result$x[, 2], Passer.Rating2.0 = Career_stats_average$Passer.Rating2.0)

pc1_pc2_plot_Passer.Rating2.0 = ggplot(as.data.frame(pca_result$x), aes(x = PC1, y = PC2, color = Career_stats_average$Passer.Rating2.0_Category)) +
  geom_point() +
  labs(x = "PC1", y = "PC2", color = "Passer.Rating2.0") +
  ggtitle("PCA: PC1 vs. PC2 seperated by passer rating") +
  scale_color_manual(values = color_mapping3)  

# Print or visualize the PCA results and plots
print(pca_result)
print(scree_plot)
print(pc1_pc2_plot_Generation)
print(pc1_pc2_plot_Games)
print(pc1_pc2_plot_Passer.Rating2.0)


# Create a biplot
biplot(pca_result, scale = 0, col = c("white", "black"))
```

